---
title: "COMA Vignette"
author: "Jeff Endelman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{COMA Vignette1}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE,comment="##",
                      dpi=150)
opts_knit$set(root.dir="~/Box Sync/Endelman/Software/COMA")
```

The starting point for analysis is the function `read_data`. The first argument is the name of a CSV input file with the marker effects and allele dosage data. The additive and dominance marker effects should correspond to a breeding value parameterization (including heterosis/inbreeding depression) and can be estimated using the `blup` command in [R/StageWise](https://github.com/jendelman/StageWise). The vignette dataset comes from the University of Wisconsin potato breeding program. There are 93 tetraploid clones genotyped at nearly 15K markers, and the marker effects are derived from a multi-trait index.

```{r}
geno.file <- system.file("vignette_data", "geno.csv", package = "COMA")
geno <- read.csv(geno.file,check.names=F)
geno[1:4,1:5]
```

The second argument to `read_data` is the name of a CSV file with a kinship matrix to control inbreeding. Kinship should be calculated based on the probability of IBD or IBS (not a G matrix). The package comes with a pedigree-based, IBD kinship matrix for the 93 clones, estimated using R package [AGHmatrix](https://cran.r-project.org/web/packages/AGHmatrix/index.html).

```{r,fig.width=4,fig.height=4}
library(COMA)
K.file <- system.file("vignette_data", "IBD.kinship.csv", package = "COMA")

#preview files
K <- as.matrix(read.csv(K.file,check.names=F,row.names=1))
K[1:3,1:3]
```

Now we are ready for `read_data`:

```{r}
data <- read_data(geno.file=geno.file,
                  kinship.file=K.file,
                  ploidy=4, matings="all",
                  standardize=TRUE, n.core=2)

head(data$parents)
head(data$matings)
```

The `matings` argument for `read_data` indicates how to create the table of possible matings. The above example used "all", which creates all unique parental combinations. When argument `standardize` is TRUE, the trait is centered and scaled so that the parental population has zero mean and unit variance. This is for convenience when interpreting the response values.

As shown above, the `parents` output contains the predicted merit (GEBV) for each individual. The `matings` output contains the predicted merit ( F1 progeny mean) for each possible mating. For dioecious species (with separate sexes), a data frame with sex information is needed as argument `sex` in `read_data`, and the column headers for the matings are "female","male" instead of "parent1","parent2". 

### OCS and OMA

The `ocs` and `oma` functions are used to determine the optimal genetic contributions. Both require the argument `parents`, which must specify the minimum and maximum contributions. In this case we used min=0 and max=1 for all parents, but this can be tailored to each parent. For `oma`, the parental merits are not used, so this column is omitted. Rather, the merit of each mating is needed, along with its min and max. Common guidance in animal breeding is to target an inbreeding rate (dF) of 0.005 to 0.01, so we begin with those values.

```{r}
parents <- data.frame(data$parents, min=0, max=1)

ocs1 <- ocs(dF=0.005, parents=parents, ploidy=4, K=data$K)
ocs2 <- ocs(dF=0.01, parents=parents, ploidy=4, K=data$K)
rbind(ocs1$response, ocs2$response)

parents2 <- data.frame(id=data$parents$id, min=0, max=1)
matings <- data.frame(data$matings, min=0, max=1)

oma1 <- oma(dF=0.005, parents=parents2, matings=matings, ploidy=4, K=data$K)
oma2 <- oma(dF=0.01, parents=parents2, matings=matings, ploidy=4, K=data$K)
rbind(oma1$response, oma2$response)
```

The above results show the expected response (merit) and Shannon diversity for the optimized parental contributions. For both OCS and OMA, higher dF leads to higher merit and lower diversity, which is an expected trade-off. The dF values returned by the software are the realized inbreeding rates, which can exceed the target due to thresholding (see below). OMA returns two realized inbreeding rates, dF1 and dF2, which correspond to the rates over the next two generations. The COMA manuscript (in preparation) provides a fuller explanation of why dF2 is needed, but numerical evidence is shown below. OMA is theoretically superior to OCS because it accommodates non-random mating, and yet the predicted merit was lower for OMA than OCS when dF=0.005. This is likely due to the approximations used in OCS, both for merit and inbreeding, and should not be interpreted as a true advantage. 

Now let's examine the optimum contributions, which are returned as `oc` for the parental contributions in both functions. The optimal mate allocations are returned as `om` by `oma`.

```{r}
kable(oma2$oc[1:5,], digits=3)
kable(oma2$om[1:5,], digits=3)
```

In the above output, many parent and mate contributions are very small (rounding to 0.000). In practice, there will be some minimum contribution that a breeding program can realize. For example, in the UW-Madison potato breeding program, approximately 20,000 progeny seedlings are generated per year per market type (e.g., round white, russet), and we prefer to work with at least 100 offspring per family. Thus, the minimum mate allocation is 100/20,000 = 0.5%, and the minimum parental contribution is half that. These practical constraints are enforced after the optimization by setting all allocations/contributions below the minimum to zero. This can generate realized inbreeding rates above the target, but the excess is small for thresholds < 1%. The following code illustrates the `min.a` and `min.c` parameters for `oma` and `ocs`, respectively.

```{r}
ocs2b <- ocs(dF=0.01, parents=parents, ploidy=4, K=data$K, min.c=0.0025)
oma2b <- oma(dF=0.01, parents=parents2, matings=matings, ploidy=4, 
             K=data$K, min.a=0.005)
ocs2b$response
oma2b$response
kable(oma2b$oc[1:5,],digits=3)
kable(oma2b$om[1:5,],digits=3)
```

Here is some code to quickly explore a range of inbreeding rates, using the `lapply` command:

```{r}
dF <- c(0.5,1,2,3)/100
ans <- lapply(as.list(dF), oma, parents=parents2, matings=matings, 
              ploidy=4, K=data$K, min.a=0.005)
response <- t(sapply(ans,"[[",1))
kable(response, digits=3)
```

The above table illustrates the significance of the dF2 parameter, which is at the upper limit even when dF1 is not.

The `plot_ribbon` command produces ribbon plots of the optimal contributions or allocations. The argument `min.c` behaves similarly as above, eliminating values below the threshold and renormalizing the proportions.

```{r,fig.width=5,fig.height=4}
oc <- lapply(ans,"[[",'oc')
names(oc) <- dF
plot_ribbon(oc)
```

```{r,fig.width=6,fig.height=4}
f <- function(z){data.frame(id=apply(z$om[,1:2],1,paste,collapse="/"),
                            value=z$om$value)}
om <- lapply(ans,f)
names(om) <- dF
plot_ribbon(om, min.c=0.01)
```

### Other Constraints

Because COMA is based on convex optimization, additional linear constraints can be incorporated. To illustrate, for the potato dataset there is additional information about two binary traits for the parental candidates. One is cytoplasmic male sterility (CMS) and the other potato virus Y resistance (PVYR), both of which were determined based on genetic markers. 

```{r}
other.file <- system.file("vignette_data", "candidates.csv", package = "COMA")
data2 <- read.csv(other.file)
table(data2[,-1])
PVYR.id <- data2$id[data2$PVYR=="Y"]
CMS.id <- data2$id[data2$CMS=="Y"]
```

The virus resistance trait is important enough that we typically require every cross to have at least one resistant parent. This can be accomplished by eliminating matings without a resistant parent (equivalent to putting the maximum allocation at 0). Furthermore, every mating must have at least one parent that is not CMS (to be the male).

```{r}
PVY.test <- matings$parent1 %in% PVYR.id | matings$parent2 %in% PVYR.id
CMS.test <- !(matings$parent1 %in% CMS.id & matings$parent2 %in% CMS.id)
matings2 <- matings[PVY.test & CMS.test,]
nrow(matings)
nrow(matings2) #eliminates 1252 possible matings

oma2c <- oma(dF=0.01, parents=parents2, matings=matings2, ploidy=4, 
             K=data$K, min.a=0.005)
oma2c$response
```

The constraints have reduced the optimal merit from 2.04 (shown earlier) to 1.77, which is to be expected. 

COMA also accepts linear equality and inequality constraints on the parental contributions. This can be used to limit the frequency of CMS clones in the mating plan, which is beneficial because every mating with a CMS parent (as the female) reduces the number of male parent candidates in the next generation. The following code shows that the total contribution of CMS parents in the previous solution was 40%.

```{r}
ix <- oma2c$oc$id %in% CMS.id
sum(oma2c$oc$value[ix])
```

Let's try limiting the total contribution to 25%. This is accomplished by adding an additional column to the `parents` data.frame, where the values represent the coefficients $\mathbf{a}$ of the contribution variables $\mathbf{y}$, and the column header is parsed to complete the equation. In this case, the header "lt0.25" represents less than or equal to 0.25:
$$ \mathbf{a}^\prime \mathbf{y} \le 0.25$$

```{r}
parents2$lt0.25 <- ifelse(parents2$id %in% CMS.id,1,0)
head(parents2)

oma2d <- oma(dF=0.01, parents=parents2, matings=matings2, ploidy=4, 
             K=data$K, min.a=0.005)
oma2d$response

ix <- oma2d$oc$id %in% CMS.id
sum(oma2d$oc$value[ix])
```

The response goes down again because of the additional constraint, which the final line verifies is met.